\bf Model Visualization Techniques for a Social Network Model

### 2017-08-21 paper.Rnw
through new section 4.2

The \texttt{RSiena} software that fits the models to data also includes a goodnes-of-fit function for examining model fit, \texttt{sienaGOF()}. This function ``assess[es] the fit of the model with respect to auxiliary statistics of networks" \citep[p.~53]{RSienamanual}. Examples of auxiliary statistics include the out- or indegree distribution on the nodes, with the option for users to input their own statistics to examine. The goodness-of-fit is evaluated as follows:
\begin{enumerate}
\item The auxiliary statistics are computed on the observed data and on $N$ simulated observations from the model. Typically, $N=1000$. 
\item The mean vector and covariance matrix of the statistics on the simulations from the model are computed.
\item The Mahalanobis distance from the observed statistics to the distribution of the simulated statistics is computed using the mean and covariance found in step 2. 
\item The Mahalanobis distrance from each of the $N$ simulations to the same distribution is computed, and the Mahalanobis distance of the observed data is compared to this distribution of distances.
\item A $p$-value is found by computing the proportion of simulated distances found in step 4 that are as large or larger than the Mahalanobis distance from the data. A SAOM is thus considered a good fit if $p$ is large.
\end{enumerate}
 %\st{XX more detail about mahalanobis distance and calculating p-value XX}

@@

\subsection{View the model in the data space}

@@

p1 <- ggplot(data = t1) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = t1$color,
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
           singletons = T) +
  labs(title = "Summary: M1") + ThemeNet +
  theme(plot.background = element_rect(color = 'black'))

@@

Another way to view the model in the data space is through simulation from the model. No single network alone simulated from a SAOM is going to look like the data or represent the model, just as no single value simulated from the standard normal distribution will look like a bell-shaped curve. It would therefore better to visualize many simulations together. For network data, one way we propose to do this is through a "summary" or "average" network drawn using the traditional node-link diagram. In Figure~\ref{fig:summnetM1}, we show a summary network created with 1,000 simulations of the second wave of the network from Model 1. \st{To make this summary network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model $M1$, for which parameters had previously been estimated. We then combine the 1,000 instances of wave 2, and count up the number of times each edge appears in a simulation. Then, we combine these 1,000 networks into a single network with edgeweight equal to the proportion of time that edge appears in our 1,000 simulations. This is the network we draw using the node-link diagram.} An edge is only drawn in the summary network if it appears in more than 5\% of the simulations (in at least 51 of the 1000 simulations), with edges that appear more frequently emphasized by thicker linewidths and a darker color. On either side of the summary network in Figure\ref{fig:summnetM1}, we show the actual data, wave 1 on the left, and wave 2 on the right. \st{We can see that the structure of the summary network is much more similar to the first network observation than to the second network observation. The simulations, however, are supposed to represent the second wave of data, which is shown on the right in Figure~\ref{fig:summnetM1}. This is an indication that our simple model, $M1$ is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.} % XXX ST - I was looking at the wrong one. I had wave == 2 instead of wave == 1 in the code. This means I was looking at the wrong simulations, so they really do look more like the first data obs than the second obs. XXX Comparing the three visualizations side-by-side shows us that the summary network of simulations of wave 2 looks more like the true wave 2 than wave 1, which is a good indication that the model is capturing the true data structure. It's not ideal, however, because there appear to be too many reciprocated ties in the summary network, far more than there are in the data. This could indicate that we should remove the reciprocity parameter from the model.

@@

The second principle of Wickham et al states that ``collections are more informative than singletons" \citep[p.~210]{modelvis}. They describe several different methods for producing such collections, but we focus on only a few here. We look at collections produced ``from exploring the space of all possible models," ``by varying model settings", ``by fitting the same model to different datasets", and ``by treating each iteration as a model" \citep[p.~210-11]{modelvis}. We also fit the same model many times, resulting in distributions of fitted parameters for one data set.

A classical problem with statistical network models generally is the lack of an ``average network" measure. Statisticians frequently rely on averages and expected values, but statistical network models, especially those as complex as SAOMs, lack an intuitive expected value measure. We could talk about expected values of parameters, but there is no way to talk about the expected value of the network based on the model. How then, can we arrive at an ``average" network? We propose to answer this question through visualization.

We these three models using \texttt{RSiena} 1,000 times each. We then looked at the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}. We can see from these distributions that the inclusion of the jumping transitive triplet parameter, $\beta_3$ is obviously affecting the distributions of the other four parameters included in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta2$. In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties. This seems problematic: it is not clear that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters.

To further investigate this relationship between the parameter values, we look at correlations between each of the parameter estimates in each model. In Figure~\ref{fig:corplots}, we examine correlations between each of pair of parameters within each model and overall.\nocite{ggally} The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute value of correlation between those two parameter values greater than 0.90 in all three models. The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2, but it is not as highly correlated with the $\beta_4$ parameter in model M3.

@@

We also explore simulations from these different models given parameter values. Using the means from the 1,000 fitted parameter values, we simulated 1,000 observations from each of our three models. We condition on the first friendship network observation, and the second and third observations are simulated from the SAOM models estimated previously. From these simulations, we first create a visualization that represents an ``average" network. To do this, we count occurences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1000 times. In Figure~\ref{fig:averagenet}, we show the first wave, the ``average'' network from the three models we fit, and the second wave. \st{Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1. Model 2, which included the transitive triplet parameter, seems to have created more group structure overall than models 1 and 3. In particular, if we look at the group of nodes $\{10,11,14\}$, we see they are very strongly connected within the three average networks, and they are completely separate from the other nodes in the true wave 2. None of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated. In Model 2, however, the ties to node 7 appear much weaker than in Model 1 or Model 2, suggesting that of the three, Model 2 is the best fit for our data.}  %XXXX ST commented out after I realized I had made the same mistake as figure 7. XX This could be an indication that, despite the highly significant nature of the parameter $\beta_3$ according to the $t$-tests in \texttt{RSiena}, the transitive triplet parameter should not be included because it does not result in a structure similar to the data.

>>>>

The \texttt{RSiena} software that fits the models to data also includes a goodnes-of-fit function for examining model fit, \texttt{sienaGOF()}. This function ``assess[es] the fit of the model with respect to auxiliary statistics of networks" \citep[p.~53]{RSienamanual}. Examples of auxiliary statistics include the out- or indegree distribution on the nodes, with the option for users to input their own statistics to examine. The goodness-of-fit is evaluated as follows:
\begin{enumerate}
\item The auxiliary statistics are computed on the observed data and on $N$ simulated observations from the model. Typically, $N=1000$. 
\item The mean vector and covariance matrix of the statistics on the simulations from the model are computed.
\item The Mahalanobis distance from the observed statistics to the distribution of the simulated statistics is computed using the mean and covariance found in step 2. 
\item The Mahalanobis distrance from each of the $N$ simulations to the same distribution is computed, and the Mahalanobis distance of the observed data is compared to this distribution of distances.
\item A $p$-value is found by computing the proportion of simulated distances found in step 4 that are as large or larger than the Mahalanobis distance from the data. A SAOM is thus considered a good fit if $p$ is large.
\end{enumerate}

@@

\subsection{View the model in the data space}\label{sec:minds}

@@

p1 <- ggplot(data = t1) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = t1$color,
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
           singletons = T) +
  labs(title = "Average: M1") + ThemeNet +
  theme(plot.background = element_rect(color = 'black'))

@@

Another way to view the model in the data space is through simulation from the model. No single network alone simulated from a SAOM is going to look like the data or represent the model, just as no single value simulated from the standard normal distribution will look like a bell-shaped curve. It would therefore better to visualize many simulations together. From a statistician's point of view, a stumbling block with statistical network models generally is the lack of an ``average network" measure. Statisticians frequently rely on averages and expected values in data analyses, but statistical network models, especially those as complex as SAOMs, lack a single, intuitive expected value measure. We could talk about expected values of parameters, but the parameters can be hard to interpret. Furthermore, there is no way to talk about the expected value of an observation from a statistical network model. How then, can we arrive at an ``average" network? We propose to answer this question through visualization. For network data, one way we propose to come up with an "average" network is through a summary network drawn using the traditional node-link diagram. In Figure~\ref{fig:summnetM1}, we show an \textit{average network} created with 1,000 simulations of the second wave of the network from Model 1. \st{To make this average network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model $M1$, for which parameters had previously been estimated. We then combine the 1,000 instances of wave 2, and count up the number of times each edge appears in a simulation. Then, we combine these 1,000 networks into a single network with edgeweight equal to the proportion of time that edge appears in our 1,000 simulations. This is the network we draw using the node-link diagram.} An edge is only drawn in the average network if it appears in more than 5\% of the simulations (in at least 51 of the 1000 simulations), with edges that appear more frequently emphasized by thicker linewidths and a darker color. On either side of the average network in Figure\ref{fig:summnetM1}, we show the actual data, wave 1 on the left, and wave 2 on the right. \st{We can see that the structure of the average network is much more similar to the first network observation than to the second network observation. The simulations, however, are supposed to represent the second wave of data, which is shown on the right in Figure~\ref{fig:summnetM1}. This is an indication that our simple model, $M1$ is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.} % XXX ST - I was looking at the wrong one. I had wave == 2 instead of wave == 1 in the code. This means I was looking at the wrong simulations, so they really do look more like the first data obs than the second obs. XXX Comparing the three visualizations side-by-side shows us that the summary network of simulations of wave 2 looks more like the true wave 2 than wave 1, which is a good indication that the model is capturing the true data structure. It's not ideal, however, because there appear to be too many reciprocated ties in the summary network, far more than there are in the data. This could indicate that we should remove the reciprocity parameter from the model.

@@

The second principle of model visualization is that ``collections are more informative than singletons" \citep[p.~210]{modelvis}. The authors describe several different methods for producing such collections, but we focus on only a few here. We look at collections produced ``from exploring the space of all possible models," ``by varying model settings", ``by fitting the same model to different datasets", and ``by treating each iteration as a model" \citep[p.~210-11]{modelvis}. Because fitting a SAOM requires MCMC methods, we also fit the same model many times, resulting in distributions of fitted parameters for one data set.

To our small frienship network, we fit three models, $M1$, $M2$, and $M3$, using \texttt{RSiena} 1,000 times each. We then looked at the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}. We can see from these distributions that the inclusion of the jumping transitive triplet parameter, $\beta_3$ is obviously affecting the distributions of the other four parameters included in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$. In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties. It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters, so we continue to explore the collection of parameter estimates.

To further investigate this potentially problematic relationship between the parameter values, we look at correlations between each of the parameter estimates in each model. In Figure~\ref{fig:corplots}, we examine correlations between each of pair of parameters within each model and overall.\nocite{ggally} The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute value of correlation between those two parameter values greater than 0.90 in all three models. The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2, but it is not as highly correlated with the $\beta_4$ parameter in model M3.

@@

We also explore simulations from these different models given parameter values. Using the mean values of from the 1,000 fitted parameter values as the parameters in our models, we simulated 1,000 observations from each of our three models. In this process, we condition on the first friendship network observation, and the second and third observations are simulated from the SAOM models with the given parameter values. From these simulations, we first create a visualization that represents an average network. To do this, we follow the same procedure as in Section~\ref{sec:minds}, counting occurences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times. As in Figure~\ref{fig:summnetM1}, edges only appear in the average network if they appear more than 5\% of the time in the simulations. In Figure~\ref{fig:averagenet}, we show the first wave, the ``average'' network from the three models we fit, and the second wave. \st{Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1. Model 2, which included the transitive triplet parameter, seems to have created a larger connected component overall than models 1 and 3. In particular, if we look at the group of nodes $\{10,11,14\}$, we see they are very strongly connected within the three average networks, and they are completely separate from the other nodes in the true wave 2. None of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated. In Model 2, however, the ties to node 7 appear much weaker than in Model 1 or Model 2, suggesting that of the three, Model 2 may be the best fit for our data.}  %XXXX ST commented out after I realized I had made the same mistake as figure 7. XX This could be an indication that, despite the highly significant nature of the parameter $\beta_3$ according to the $t$-tests in \texttt{RSiena}, the transitive triplet parameter should not be included because it does not result in a structure similar to the data.
